Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-11-03 demystify.tex
updates

Students from top schools with strong research programs and experience might already know most of the information given here.
My goal is thus to level the playing field by providing info that is not readily available to less privileged students. 
I hope to encourage more students with such backgrounds to apply to CS PhD programs in the US.

Another goal of this document is to explain the \emph{``why''} behind the admission process, e.g., why LoRs are important, why you should not draft your own LoR, why you should contact professors, etc. Understanding the reason and mindset of the admission committee and profs. can help you prepare better.

While this document may not fully meet the needs of students aiming for top-10 schools (\autoref{sec:selecting-ranking-schools} and \autoref{sec:ranking}), which typically have more stringent requirements (e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU), it should be valuable for those targeting other reputable R1 institutions. It is certainly useful for students aiming for Mason!

%\begin{commentbox}[Vu:]
%    One of the reasons I created this document is that my colleagues at Mason are interested in recruiting Vietnamese students and are surprised when seeing very few applications in Vietnam (see \autoref{sec:ack}). Each year our CS PhD program receives 500+ PhD applications, most of which are international but only 5--6 are from Vietnam.
%  \end{commentbox}

@@

Typically, it takes 5--7 years for CS PhD in the US.  This can be longer than CS PhDs in other countries (\autoref{sec:non-us-differences}), which might require MS (\autoref{sec:phd-vs-ms}) first.

@@

Given the high number of quality applicants and a limited number of spots, in addition to having a good application profile, you want to show something that makes you \textbf{stand out}.  For example, even if you do not have research experience, you can talk about your personal projects, as long as they can help show you can do research. For example, if you have an open-source project that has lots of stars in Github, then you should mention it. If you often write technical, research-like blogs with many viewers, talk about them too.

There are other things you might not think are important but can make you stand out. For example, if you have a strong background in a non-CS field but can be integrated with CS, e.g., you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS, do talk about it. Are you a female or a minority in CS (\autoref{sec:urm})? Do you participate in outreach activities that help increase diversity and inclusion in CS? Diversity is highly valued in CS programs in the US, and experiences in these areas can make you stand out and get noticed from reviewers.

@@

@@

\section{Differences between PhD and MS}\label{sec:phd-vs-ms}

>>>>

Students from top schools with strong research cultures and experience might already know most of the information given here.
My goal is thus to level the playing field by providing information that is not readily available to less privileged students. 
I hope to encourage more students with such backgrounds to apply to CS PhD programs in the US.
Another goal of this document is to explain the \emph{``why''} behind the admission process, e.g., why LoRs are important, why you should not draft your own LoR, why you should contact professors, etc.

While this document may not fully meet the needs of students aiming for top-10 schools (\autoref{sec:ranking}), which typically have more stringent requirements (e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU), it should be valuable for those targeting other reputable R1 institutions. It is certainly useful for students aiming for Mason!

\begin{commentbox}[Vu:]
    One of the reasons I created this document is that my colleagues at Mason are interested in recruiting Vietnamese students and are surprised when seeing very few applications in Vietnam (see \autoref{sec:ack}). Each year our CS PhD program receives 500+ PhD applications, most of which are international but only 5--6 are from Vietnam. In general the number of CS PhD applications from Vietnam to US universities is very low and more would be welcome.
  \end{commentbox}

@@

Typically, it takes 5--7 years for CS PhD in the US.  This can be longer than CS PhDs in other countries (\autoref{sec:msrequirement}), which might require MS first.

@@

Given the high number of quality applicants and a limited number of spots, in addition to having a good application profile, you want to show something that makes you \textbf{stand out}.  For example, even if you do not have research experience, you can talk about your personal projects, as long as they show you can do research. For example, if you have an open-source project that has lots of stars in Github, mention it. If you often write technical, research-like blogs with many viewers, talk about them too.

@@

There are other things you might not think are important but can make you stand out. For example, if you have a strong background in a non-CS field but can be integrated with CS, e.g., you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS, do talk about it. Are you a female or a minority in CS? Do you participate in outreach activities that help increase diversity and inclusion in CS? Diversity is highly valued in CS programs in the US (\autoref{sec:urm}), and experiences in these areas can make you stand out and get noticed from reviewers.

@@

\section{Differences between PhD and MS}
